### 2018-03-12 submission/manuscript.Rmd
good read through upto results

**Results.** We quantified the odds ratios for individual bacterial genera that were associated with an individual having tumors relative to a normal colon. Among the stool samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_OR_stool$measure)` genera with significant odds ratios (ORs) associated with carcinoma. Similarly, among the tissue samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` genera with significant odds ratios associated with carcinoma. Among the significant odds ratios, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`. Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using only the taxa that had significant ORs, using the entire collection of taxa found in each study, and using operational taxonomic units defined based on a 97% similarity threshold. All training approaches yielded similar classification success as measured using the Area Under the Curve. The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from stool or tissue.

@@

Recently, there has been an intense focus on identifying microbiota-based biomarker yielding a seemingly endless number of candidate taxa. Some studies point towards mouth-associated genera such as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that are enriched in people with carcinomas [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Other studies have identified members of *Akkermansia*, *Bacteroides*, *Enterococcus*, *Escherichia*, *Klebsiella*, *Mogibacterium*, *Streptococcus*, and *Providencia* are also associated with carciomas  [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. Additionally, *Roseburia* has been found in some studies to be more abundant in people with tumors but in other studies it has been found to be either less abundant or no different than what is found in subjects with normal colons  [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014]. There are strong results from tissue culture and murine models that Fusobacterium nucleatum, pks-positive strains of Escherichia coli, Streptococcus gallolyticus, and an entertoxin-producing strain of Bacteroides fragilis are important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. These results point to a causative role for the microbiota in CRC pathogenesis as well as their potential as diagnostic biomarkers.

@@

Confirming some of these previous findings, a recent meta-analysis found that 16S rRNA gene sequences from members of the *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017]. Contrary to previous studies they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014]. In addition, they found 16S rRNA gene sequences from members of *Haemophilus*, *Methanosphaera*, *Prevotella*, *Succinovibrio* were enriched in patients with adenoma and *Pantoea* were enriched in patients with carcinomas. Although this meta-analysis was helpful for distilling a large number of possible biomarkers, the aggregate number of samples included in the analysis (n = 509) was smaller than several larger case-control studies that have since been published [@baxter_microbiota-based_2016; @hale_shifts_2017]

Here we provide an updated meta-analysis using 16S rRNA gene sequence data from both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) from 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2]. We expand both the breadth and scope of the previous meta-analysis to investigate whether biomarkers describing the bacterial community or specific members of the community can more accurately classify patients as having adenoma or carcinoma. Our results suggest that the bacterial community changes as disease severity worsens and that that a subset of the microbial community can be used to diagnose the presence of carcinoma.

@@

***Lower bacterial diversity is associated with increased odds ratio (OR) of tumors.***
We first assessed whether variation in broad community metrics like total number of operational taxonomic units (OTUs) (i.e. richness), the evenness of their abundance, and the overall diversity was associated with disease stage after controlling for study and variable region differences. In stool, there was a significant decrease in both evenness and diversity as disease severity progressed from normal to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1]; there was not a significant difference for richness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`). We next tested whether the decrease in these community metrics translated into significant ORs for having an adenoma or carcinoma. For fecal samples, the ORs for richness were not significantly greater than 1.0 for adenoma or carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "sobs")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A]. The ORs for evenness were significantly higher than 1.0 for adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and carcinoma (OR =  `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2B]. The ORs for diversity were only significantly greater than 1.0 for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`), but not for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2C]. Although these OR are significantly greater than 1.0, it is doubtful that these are clinically meaningful values.

@@

***Disease progression is associated with community-wide changes in composition and abundance.*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages. We identified significant bacterial community differences in the stool of patients with adenomas relative to those with normal colons in `r length(as.data.frame(filter(adn_beta_stool, pvalue < 0.05))[, "study"])` of `r length(as.data.frame(adn_beta_stool)[, "study"])` studies and in patients with carcinomas relative to those with normal colons in 6 of 7 studies (PERMANOVA; P-value < 0.05) [Table S2]. Similar to the analyses using stool samples, there were significant differences in bacterial community structure between subjects with normal colons and those with adenoma (1 of 2 studies) and carcinoma (1 of 3 studies) [Table S2]. For studies that used matched samples no differences in bacterial community structures were observed [Table S2]. Combined, these results indicate that there consistent and significant community-wide changes in the fecal community structure of subjects with carcinomas. However, the signal observed in subjects with adenomas or when using tissue samples was not as consistent. This is likely due to a smaller effect size or the relatively small sample sizes among the studies that characterized the tissue microbiota.

@@

Next we identified those taxa were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas. No taxa had a significant OR for the presence of adenomas when we used data collected from stool or tissue samples (Table S3 & S4). In contrast, `r length(crc_OR_stool$measure)` taxa had significant ORs for the presence of carcinomas using data from stool samples. Of these, 4 are commonly associated with the oral cavity: *Fusobacterium* (OR = `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`)), *Parvimonas* (OR = `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_ub"], digits = 2))`)), *Porphyromonas* (OR = `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2))`)), and *Peptostreptococcus* (OR = `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_ub"], digits = 2))`)) [Table S3]. The other 4 were *Clostridium XI* (OR = `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), *Enterobacteriaceae*  (OR = `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), *Escherichia* (OR = `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and *Ruminococcus* (OR = `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)). Among the data collected from tissue samples, only unmatched carcinoma samples had taxa with a significant OR. Those included *Dorea* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2))`)), *Blautia* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2))`)), and *Weissella* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2))`)). Mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4]. For example, *Fusobacterium* had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg-corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4]. It is interesting to note that *Ruminococcus* and members of *Clostridium group XI* in stool and *Dorea* and *Blautia* in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas. Overall, there was no overlap in the taxa with significant OR between stool and tissue samples.

***Individual taxa with a significant OR do a poor job of differentiating subjects with normal colons and those with carcinoma.***
 We next asked whether those taxa that had a significant OR associated with having a normal colon or carcinomas could be used individually, to classify subjects as having a normal colon or carcinomas. OR values were caluclated based on whether the relative abundance for a taxon in a subject was above or below the median relative abundance for that taxon across all subjects in a study. To measure the ability of these taxa to classify individuals we instead generated receiver operator characteristic (ROC) curves for each taxon in each study and calculated the area under the curve (AUC). This allowed us to use a more fluid relative abundance threshold for defining disease status. Using data from stool samples, the `r length(crc_OR_stool$measure)` taxa did no better at classifying the subjects than one would expect by chance (i.e. AUC=0.50) [Figure 3A]. The taxa that performed the best included *Clostridium XI*, *Ruminococcus*, and *Escherichia*, however, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`. Likewise, in unmatched tissue samples the `r length(crc_OR_stool$measure)` taxa with significant ORs taxa were marginally better than one would expect by chance [Figure 3B]. The relative abundance of Dorea was the best predictor of carcinomas and its median AUC was only `r round(max(filter(ind_crc_combined_unmatched_auc, taxa == "Dorea")[, "auc"]), digits = 3)`. These results suggest that although these taxa are associated with a decreased or increased OR for the presences of carcinomas, individually, they do a poor job of classifying a subject’s disease status.

>>>>

**Results.** We quantified the odds ratios for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon. Among the stool samples, there were no taxa that had a significant odds ratio associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant odds ratios (ORs) associated with carcinoma. Similarly, among the tissue samples, there were no taxa that had a significant odds ratio associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` taxa with significant odds ratios associated with carcinoma. Among the significant odds ratios, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`. Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using only the taxa that had significant ORs, using the entire collection of taxa found in each study, and using operational taxonomic units defined based on a 97% similarity threshold. All training approaches yielded similar classification success as measured using the Area Under the Curve. The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from stool or tissue.

@@

Recently, there has been an intense focus on identifying microbiota-based biomarkers yielding a seemingly endless number of candidate taxa. Some studies point towards mouth-associated genera such as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that are enriched in people with carcinomas [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Other studies have identified members of *Akkermansia*, *Bacteroides*, *Enterococcus*, *Escherichia*, *Klebsiella*, *Mogibacterium*, *Streptococcus*, and *Providencia* are also associated with carciomas  [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. Additionally, *Roseburia* has been found in some studies to be more abundant in people with tumors but in other studies it has been found to be either less abundant or no different than what is found in subjects with normal colons  [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014]. There are strong results from tissue culture and murine models that *Fusobacterium nucleatum*, pks-positive strains of *Escherichia coli*, *Streptococcus gallolyticus*, and an entertoxin-producing strain of *Bacteroides fragilis* are important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. These results point to a causative role for the microbiota in CRC pathogenesis as well as their potential as diagnostic biomarkers.

@@

Confirming some of these previous findings, a recent meta-analysis found that 16S rRNA gene sequences from members of  *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017]. Contrary to previous studies they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014]. In addition, they found 16S rRNA gene sequences from members of *Haemophilus*, *Methanosphaera*, *Prevotella*, *Succinovibrio* were enriched in patients with adenoma and *Pantoea* were enriched in patients with carcinomas. Although this meta-analysis was helpful for distilling a large number of possible biomarkers, the aggregate number of samples included in the analysis (n = 509) was smaller than several larger case-control studies that have since been published [@baxter_microbiota-based_2016; @hale_shifts_2017]

Here we provide an updated meta-analysis using 16S rRNA gene sequence data from both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) from 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2]. We expand both the breadth and scope of the previous meta-analysis to investigate whether biomarkers describing the bacterial community or specific members of the community can more accurately classify patients as having adenoma or carcinoma. Our results suggest that the bacterial community changes as disease severity worsens and that a subset of the microbial community can be used to diagnose the presence of carcinoma.

@@

***Lower bacterial diversity is associated with increased odds ratio (OR) of tumors.***
We first assessed whether variation in broad community metrics like total number of operational taxonomic units (OTUs) (i.e. richness), the evenness of their abundance, and the overall diversity was associated with disease stage after controlling for study and variable region differences. In stool, there was a significant decrease in both evenness and diversity as disease severity progressed from normal to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1]; there was no significant difference for richness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`). We next tested whether the decrease in these community metrics translated into significant ORs for having an adenoma or carcinoma. For fecal samples, the ORs for richness were not significantly greater than 1.0 for adenoma or carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "sobs")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A]. The ORs for evenness were significantly higher than 1.0 for adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)(95% Confidence Interval), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and carcinoma (OR =  `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2B]. The ORs for diversity were only significantly greater than 1.0 for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`), but not for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2C]. Although these OR are significantly greater than 1.0, it is doubtful that these are clinically meaningful values.

@@

***Disease progression is associated with community-wide changes in composition and abundance.*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages. We identified significant bacterial community differences in the stool of patients with adenomas relative to those with normal colons in `r length(as.data.frame(filter(adn_beta_stool, pvalue < 0.05))[, "study"])` of `r length(as.data.frame(adn_beta_stool)[, "study"])` studies and in patients with carcinomas relative to those with normal colons in 6 of 7 studies (PERMANOVA; P-value < 0.05) [Table S2]. Similar to the analyses using stool samples, there were significant differences in bacterial community structure between subjects with normal colons and those with adenoma (1 of 2 studies) and carcinoma (1 of 3 studies) [Table S2]. For studies that used matched samples no differences in bacterial community structures were observed [Table S2]. Combined, these results indicate that there are consistent and significant community-wide changes in the fecal community structure of subjects with carcinomas. However, the signal observed in subjects with adenomas or when using tissue samples was not as consistent. This is likely due to a smaller effect size or the relatively small sample sizes among the studies that characterized the tissue microbiota.

@@

Next we identified those taxa that were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas. No taxa had a significant OR for the presence of adenomas when we used data collected from stool or tissue samples (Table S3 & S4). In contrast, `r length(crc_OR_stool$measure)` taxa had significant ORs for the presence of carcinomas using data from stool samples. Of these, 4 are commonly associated with the oral cavity: *Fusobacterium* (OR = `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`)), *Parvimonas* (OR = `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_ub"], digits = 2))`)), *Porphyromonas* (OR = `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2))`)), and *Peptostreptococcus* (OR = `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_ub"], digits = 2))`)) [Table S3]. The other 4 were *Clostridium XI* (OR = `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), *Enterobacteriaceae*  (OR = `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), *Escherichia* (OR = `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and *Ruminococcus* (OR = `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)). Among the data collected from tissue samples, only unmatched carcinoma samples had taxa with a significant OR. Those included *Dorea* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2))`)), *Blautia* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2))`)), and *Weissella* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2))`)). Mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4]. For example, *Fusobacterium* had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4]. It is interesting to note that *Ruminococcus* and members of *Clostridium XI* in stool and *Dorea* and *Blautia* in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas. Overall, there was no overlap in the taxa with significant OR between stool and tissue samples.

***Individual taxa with a significant OR do a poor job of differentiating subjects with normal colons and those with carcinoma.***
 We next asked whether those taxa that had a significant OR associated with having a normal colon or carcinomas could be used individually, to classify subjects as having a normal colon or carcinomas. OR values were caluclated based on whether the relative abundance for a taxon in a subject was above or below the median relative abundance for that taxon across all subjects in a study. To measure the ability of these taxa to classify individuals we instead generated receiver operator characteristic (ROC) curves for each taxon in each study and calculated the area under the curve (AUC). This allowed us to use a more fluid relative abundance threshold for defining disease status. Using data from stool samples, the `r length(crc_OR_stool$measure)` taxa did no better at classifying the subjects than one would expect by chance (i.e. AUC=0.50) [Figure 3A]. The taxa that performed the best included *Clostridium XI*, *Ruminococcus*, and *Escherichia*, however, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`. Likewise, in unmatched tissue samples the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant OR taxa were marginally better than one would expect by chance [Figure 3B]. The relative abundance of *Dorea* was the best predictor of carcinomas and its median AUC was only `r round(max(filter(ind_crc_combined_unmatched_auc, taxa == "Dorea")[, "auc"]), digits = 3)`. These results suggest that although these taxa are associated with a decreased or increased OR for the presences of carcinomas, individually, they do a poor job of classifying a subject’s disease status.

***Combined taxa model classifies subjects better than using individual taxa.*** 
Instead of attempting to classify subjects based on individual taxa, next we combined information from the individual taxa and evaluated the ability to classify a subject’s disease status using Random Forest models. For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`). For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S2]. Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study. Within the significant OR taxa only models the most important taxa across study for stool were *Ruminococcus* and *Clostridium XI* based on mean decrease in accuracy [Figure 4A]. Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for *Blautia* and *Weissella* (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).  For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S3]. For the significant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the most important based on mean decrease in accuracy [Figure 4B]. Pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.
